---
title: | 
       | Statistical Inference - Final Project: Part 2
       | Basic Inferential Data Analysis with R
       
author: "Anastasia Thöns"
date: "December 26, 2016"
output: pdf_document
---

#Introduction

In the project `ToothGrowth` data set is used - *The Effect of Vitamin C on Tooth Growth in Guinea Pigs* (Ref.: [ETH Zürich](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html)). 
The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods - orange juice or ascorbic acid (a form of vitamin C and coded as VC).
The data `ToothGrowth` is a data frame with 60 observations on 3 variables:

> [,1]	len -	numeric	Tooth length 

> [,2]	supp -	factor	Supplement type (VC or OJ). 

> [,3]	dose -	numeric	Dose in milligrams/day 

and it is a part of the library `datasets`.


###Source
* C. I. Bliss (1952) The Statistics of Bioassay. Academic Press.

###References
* McNeil, D. R. (1977) Interactive Data Analysis. New York: Wiley.

#Implementation in R
The following R libraries are used:
```{r, echo=TRUE}
library(datasets)
library(data.table)
library(ggplot2)
```

Load the `ToothGrowth` data as a `data.table` structure and give descriptive names. 
```{r, echo=TRUE}
data(ToothGrowth)
ToothGrowth <- data.table(ToothGrowth) 
names(ToothGrowth) <- c("tooth.length", "supplement.type", "dose")
```

Odontoblast length in the data set varies in the range $4.2--33.9$ in microns with the mean value $\mu=18.8$ and the variance $\sigma^2 = 58.5$.
```{r}
mean(ToothGrowth$tooth.length)
var(ToothGrowth$tooth.length)
```

Basic summary of the data set can be found in the following table:
```{r pander, echo=FALSE}
library(pander)
panderOptions("digits", 2)
pander(summary(ToothGrowth))
```

Basic scatterplot of the data set divided by the supplement type:

```{r, echo=FALSE, fig.width=8, fig.height=5}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
p1 <- ggplot(ToothGrowth, aes(x = dose, y = tooth.length, color=supplement.type))
p1 + geom_boxplot() + geom_point(size = 2, shape=1) +
      scale_color_manual(name   = "Supplement type", 
                         values = c("orange", "blue"), 
                         labels = c("orange juice", "ascorbic acid")) +
      labs(title="The Effect of Vitamin C on Tooth Growth in Guinea Pigs", 
           x="Dose in milligrams/day", y="Tooth length") 
```

Conditioning plot of the data sets:

```{r, echo=FALSE, fig.width=8, fig.height=5}
require(graphics)
coplot(tooth.length ~ dose | supplement.type, data = ToothGrowth, panel = panel.smooth,
       ylab="Tooth length",
       xlab = "ToothGrowth data: length vs dose, given type of supplement")
```

Looking at the graphics we can assume that the effect of vitamin C on tooth growth is proportional to the dose and also appears to be higher with orange juice than with ascorbic acid at least at lower dose (< 2 mg/day). In the following we will provide some statistical inference analysis to confirm of reject the statement.

#Confidence Intervals
We now analyse the correlation between the tooth length and the supplement type by Welch's unpaired t-test:
```{r, echo=TRUE}
t.test(tooth.length ~ supplement.type, paired = FALSE, ar.equal = TRUE,  data = ToothGrowth)
```
The  95 percent confidence interval $[-0.17, 7.57]$ is not fully positive. 

Let us now perform the hypothesis test. The null hypothesis is that there is no correlation between the tooth grows and the delivery method, i.e. the means are the same. In the previous we can see already that the mean in case of orange juice supplement is higher then in case of ascorbic acid. The values $Z$ and $Z_\alpha$ can be found as following:
```{r, echo=TRUE}
g1 <- ToothGrowth$tooth.length[ToothGrowth$supplement.type=="VC"]
g2 <- ToothGrowth$tooth.length[ToothGrowth$supplement.type=="OJ"]
difference <- g2 - g1
mn <- mean(difference)
s <- sd(difference)
n <- 30
Z <- sqrt(n)*(mn - 0)/s
Z
Z.alpha <- qnorm(0.95)
Z.alpha
```
And so we can see that $Z= 3.3 > 1.64$ and we reject the null hypothesis.

#Conclusions
From the performed statistical analysis on the data `ToothGrowth` we can conclude the following:

* The tooth length in tested guinea pigs is proportional to the dose supplemented of vitamin C based on exploratory data analyses and regression line.

* Based on hypothesis testing we conclude that the natural supplement type, i.e. orange juice, has a stronger affect on the tooth growth then the synthetic one, i.g. ascorbic acid.
